---
title: "HT-29 CRISPR-Cas9 screen assessment"
author: "Raffaele M. Iannuzzi, Francesco Iorio"
date: "`r Sys.Date()`"
output: pdf_document
pdf_document: 
  citation_package: biblatex
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::write_bib(x = c("knitr", "rmarkdown") , file = "test.bib")
```

## Introduction

This document guides the user through (and shows results from) the execution of functions included in the \textit{HT29benchmark} R package. The aim of these functions is to assess novel experimental pipelines for genome-wide CRISPR-Cas9 screens upon the execution of a single calibration screen of the HT-29 human colon cancer cell line (<https://www.cellosaurus.org/CVCL_0320>) employing a commercially available genome-wide library of single guide RNAs (the Sanger library) [@Tzelepis2016-vb] (AddGene: 67989) and setting described in @Behan2019-wu.\
The user provided calibration-screen data is evaluated through different metrics. Outcomes are then contrasted with those obtained when applying the same metrics to a high-quality dataset containing results from screening the HT-29 in multiple batches with the Sanger library (the HT-29 reference dataset).\

Portions of this document can be commented to generated a quality-control assessment report.\

The HT29benchmark R package is available at: <https://github.com/DepMap-Analytics/HT29benchmark>; with user reference manual available at <https://github.com/DepMap-Analytics/HT29benchmark/blob/master/HT29benchmark.pdf>. A manuscript describing the HT-29 reference dataset and the \textit{HT29benchmark} package is currently under review. The HT-29 reference dataset can be downaloded through a dedicated function of the \textit{HT29benchmark} package, as well is available on FigShare (@Behan_M2022-ks).

## Environment preparation and data retrival

The following code chunks load all required libraries installing the missing ones from Bioconductor and CRAN:\

```{r eval=FALSE, include=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

toInstall<- setdiff(c("topGO", "clusterProfiler","org.Hs.eg.db","enrichplot"),
                    rownames(installed.packages()))

BiocManager::install(toInstall)

toInstall<- setdiff(c("VennDiagram", "data.table", "KernSmooth"),
                    rownames(installed.packages()))

install.packages(toInstall)
```

```{r eval=TRUE, include=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
# Needed for the analysis
library(CRISPRcleanR)
library(HT29benchmark)

# Needed for the report
library(data.table)
library(VennDiagram)
library(clusterProfiler)
library(enrichplot) 
library(org.Hs.eg.db)
library(topGO)
library(RColorBrewer)
```

The following code creates a directory (HT29R_resFolder) in the local folder, in which the HT-29 reference dataset (sgRNA depletion fold changes or sgRNA counts) are downloaded. A subdirectory is also created (USER) and used to save plots and other figures (if the saveToFig parameter of the HT29benchmark functions is set to 'TRUE').\

```{r warning=FALSE}
dir.create('~/HT29R_resFolder/')
tmpDir <- path.expand('~/HT29R_resFolder/')
dir.create(paste(tmpDir, "USER/",sep=""))
resultsDir <- paste(tmpDir,"USER/", sep="")
```

The following code downloads the HT-29 reference dataset (i.e. sgRNA depletion log fold-changes from high-quality HT-29 screens) and stores it in the HT29_resFolder directory.\

```{r eval=FALSE, echo=TRUE, include=TRUE}
HT29R.downloadRefData(destFolder = tmpDir, whatToDownload = 'FCs')
```

The code chunk below downloads a demo screen in the 'HT29R_resFolder'. This encompasses data from a 6-replicates mid-quality screen of the HT-29 cell line, employing the Sanger library [@Tzelepis2016-vb] and setting described in @Behan2019-wu (see manuscript for more details), and it is available on FigShare (@Behan_M2022-ks).\

IMPORTANT: The following code should not be executed to perform the analysis of real user-data. See further code chunk.

```{r }
URL <- 'https://figshare.com/ndownloader/files/36658530?private_link=5b2a579791c47a417474'
download.file(URL, destfile = paste0(tmpDir, '/Example_UserScreen.tsv'))
userDataPATH <- paste0(tmpDir, '/Example_UserScreen.tsv')
```
IMPORTANT: The following code should be uncommented and executed for the analysis of real user-data. The path to the real user dataset should be provided. This should be string specifying the path to a tsv file containing the raw sgRNA counts of a
calibration screen of the HT-29 cell line performed with the Sanger library [@Tzelepis2016-vb] and experimental setting described in @Behan2019-wu\
The format of this file should be as it follows. The file should be tab delimited, it should contain one row per sgRNA and the following columns/headers:

  + sgRNA: column with alphanumerical identifiers of the sgRNA under consideration;
  + gene: column with HGNC symbols of the genes targeted by the sgRNA under consideration;

followed by the columns containing the sgRNAs' counts for the controls and columns for library transfected samples (one column per replicate).

```{r echo=TRUE}
## userDataPATH <- 'PATH/TO/REAL-USER-DATA/userdata.tsv'
```

## User data normalisation and computation of depletion log fold-changes

The following code normalises user-data sgRNA counts, computes log fold-changes and stores results.

```{r fig.align='center', fig.height=6, fig.width=12}
data('KY_Library_v1.0')
expData <- ccr.NormfoldChanges(filename = userDataPATH,
                               Dframe = NULL,
                               min_reads = 30, 
                               EXPname = "User-Screen", 
                               libraryAnnotation = KY_Library_v1.0,
                               saveToFig = FALSE,
                               outdir = resultsDir,
                               display = TRUE)
```

# References
