---
title: "HT-29 CRISPR-Cas9 screen assessment"
author: "Raffaele M. Iannuzzi, Francesco Iorio"
date: "`r Sys.Date()`"
output: pdf_document
pdf_document: 
  citation_package: biblatex
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::write_bib(x = c("knitr", "rmarkdown") , file = "test.bib")
```

## Introduction

This document guides the user through (and shows results from) the execution of functions included in the \textit{HT29benchmark} R package. The aim of these functions is to assess novel experimental pipelines for genome-wide CRISPR-Cas9 screens upon the execution of a single calibration screen of the HT-29 human colon cancer cell line (<https://www.cellosaurus.org/CVCL_0320>) employing a commercially available genome-wide library of single guide RNAs (the Sanger library) (@Tzelepis2016-vb) (AddGene: 67989) and setting described in @Behan2019-wu.\

The user-provided calibration-screen data is evaluated through different metrics. Outcomes are then contrasted with those obtained when applying the same metrics to a high-quality dataset containing results from screening the HT-29 in multiple batches with the Sanger library (the HT-29 reference dataset).\

Portions of and text and code in this document can be commented/removed to generated a quality-control assessment report.\

The HT29benchmark R package is available at: <https://github.com/DepMap-Analytics/HT29benchmark>; with user reference manual available at <https://github.com/DepMap-Analytics/HT29benchmark/blob/master/HT29benchmark.pdf>. A manuscript describing the HT-29 reference dataset and the \textit{HT29benchmark} package is currently under review. The HT-29 reference dataset can be downaloded through a dedicated function of the \textit{HT29benchmark} package, as well is available on FigShare (@Behan_M2022-ks).

## Environment preparation and data retrival

The following code chunks load all required libraries installing the missing ones from Bioconductor and CRAN:\

```{r eval=FALSE, include=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

toInstall<- setdiff(c("topGO", "clusterProfiler","org.Hs.eg.db","enrichplot"),
                    rownames(installed.packages()))

BiocManager::install(toInstall)

toInstall<- setdiff(c("VennDiagram", "data.table", "KernSmooth"),
                    rownames(installed.packages()))

install.packages(toInstall)
```

```{r eval=TRUE, include=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
# Needed for the analysis
library(CRISPRcleanR)
library(HT29benchmark)

# Needed for the report
library(data.table)
library(VennDiagram)
library(clusterProfiler)
library(enrichplot) 
library(org.Hs.eg.db)
library(topGO)
library(RColorBrewer)
```

The following code creates a directory (HT29R_resFolder) in the local folder, in which the HT-29 reference dataset (sgRNA depletion fold changes or sgRNA counts) are downloaded. A subdirectory is also created (USER) and used to save plots and other figures (if the saveToFig parameter of the HT29benchmark functions is set to 'TRUE').\

```{r warning=FALSE}
dir.create('~/HT29R_resFolder/')
tmpDir <- path.expand('~/HT29R_resFolder/')
dir.create(paste(tmpDir, "USER/",sep=""))
resultsDir <- paste(tmpDir,"USER/", sep="")
```

The following code downloads the HT-29 reference dataset (i.e. sgRNA depletion log fold-changes from high-quality HT-29 screens) and stores it in the HT29_resFolder directory.\

```{r eval=FALSE, echo=TRUE, include=TRUE}
HT29R.downloadRefData(destFolder = tmpDir, whatToDownload = 'FCs')
```

The code chunk below downloads a demo screen in the 'HT29R_resFolder'. This encompasses data from a 6-replicates mid-quality screen of the HT-29 cell line, employing the Sanger library [@Tzelepis2016-vb] and setting described in @Behan2019-wu (see manuscript for more details), and it is available on FigShare (@Behan_M2022-ks).\

IMPORTANT: The following code should not be executed to perform the analysis of real user-data. See further code chunk.

```{r eval=FALSE}
URL <- 'https://figshare.com/ndownloader/files/36658530?private_link=5b2a579791c47a417474'
download.file(URL, destfile = paste0(tmpDir, '/Example_UserScreen.tsv'))
```

```{r eval=TRUE}
userDataPATH <- paste0(tmpDir, '/Example_UserScreen.tsv')
```

IMPORTANT: The following code should be uncommented and executed for the analysis of real user-data. The path to the real user dataset should be provided. This should be string specifying the path to a tsv file containing the raw sgRNA counts of a
calibration screen of the HT-29 cell line performed with the Sanger library [@Tzelepis2016-vb] and experimental setting described in @Behan2019-wu\
The format of this file should be as it follows. The file should be tab delimited, it should contain one row per sgRNA and the following columns/headers:

  + sgRNA: column with alphanumerical identifiers of the sgRNA under consideration;
  + gene: column with HGNC symbols of the genes targeted by the sgRNA under consideration;

followed by the columns containing the sgRNAs' counts for the controls and columns for library transfected samples (one column per replicate).

```{r echo=TRUE}
## userDataPATH <- 'PATH/TO/REAL-USER-DATA/userdata.tsv'
```

The following code download sgRNA depletion fold-changes from the HT-29 reference dataset and stores them into the HT29_resFolder directory.
Additionally, individual reference screens file-names are stored in the ref_fn variable\

```{r eval=FALSE, echo=TRUE}
HT29R.downloadRefData(destFolder = tmpDir, whatToDownload = 'FCs')
```

```{r echo=TRUE}
ref_fn <- dir(tmpDir)
ref_fn <- grep('_foldChanges.Rdata',ref_fn,value=TRUE)
```

## User data normalisation and computation of depletion log fold-changes

The following code normalises user-data sgRNA counts, computes log fold-changes, stores results in the USER subfolder and produces the plots below with distributions of sgRNA counts pre-/post-normalisation and depletion log fold-changes across samples. The ccr.NormfoldChanges function is from our previously published CRISPRcleanR package (@Iorio2018-qt). Results are also stored in the UserData variable. This is a list containing two data frames respectively including the normalised sgRNAs' counts (norm_counts) and the sgRNAs' log fold changes (logFCs). First two columns in these data frames contain sgRNAs' identifiers and HGNC symbols of targete gene, respectively.

```{r fig.align='center', fig.height=6, fig.width=12}
data('KY_Library_v1.0')
UserData <- ccr.NormfoldChanges(filename = userDataPATH,
                               Dframe = NULL,
                               min_reads = 30, 
                               EXPname = "User-Screen", 
                               libraryAnnotation = KY_Library_v1.0,
                               saveToFig = FALSE,
                               outdir = resultsDir,
                               display = TRUE)
```

## Inspection of sgRNA depletion log fold-change distributions

The following code allows visualising (through dedicated plots) distributions of sgRNA log fold-changes from the HT-29 reference dataset (across screens) as well as user screen data.\

Average values and confidence intervals of these distribution are also visualised and can be compared.\ 

```{r echo=TRUE, fig.align='center', fig.height=5, fig.width=10}
HT29R.FCdistributions(refDataDir = tmpDir,
                      resDir = resultsDir, 
                      userFCs = UserData$logFCs,
                      stats = TRUE,
                      saveToFig = FALSE,
                      display = TRUE)
```

## Intra-screen reproducibility

The following code evaluates and compares inter-screen replicate similarities for the HT-29 reference screen, as well as for the user-provided screen.
Particularly, it shows Pearson’s correlation scores computed between sgRNA depletion log fold-change 
profiles of replicates for each of the six HT-29 reference screens (blue dots), and between replicates of the user-provided screen (pink dots).
These correlation scores are computed considering only depletion log fold-changes of highly
reproducible/informative sgRNAs only (defined as in @Behan2019-wu). These scores are compared with correlation scores from comparing replicates of the same screen across > 200 cell lines (from Project Score (@Behan2019-wu, @Dwane2021-lx), in green), those obtained from comparing each possible pair of replicates, regardless the screen (again from Project Score (@Behan2019-wu, @Dwane2021-lx), in gray) and a quality threshold, derived from these two distributions, as defined in @Behan2019-wu.


```{r echo=TRUE, fig.align='center', fig.height=8, fig.width=9, warning=FALSE}
HT29R.evaluateReps(refDataDir = tmpDir,
                   resDir = resultsDir,
                   userFCs = UserData$logFCs, 
                   geneLevel = FALSE, 
                   display = TRUE, 
                   saveToFig = FALSE)
```
The following code chunk performs the same comparisons of the previous one, but this time at the gene level, collapsing sgRNA depletion log fold-changes by averaging on a targeted gene basis.

```{r echo=TRUE, fig.align='center', fig.height=8, fig.width=9, warning=FALSE}
HT29R.evaluateReps(refDataDir = tmpDir,
                   resDir = resultsDir,
                   userFCs = UserData$logFCs, 
                   geneLevel = TRUE, 
                   display = TRUE, 
                   saveToFig = FALSE)
```

## Inter-screen similarity

Text

```{r echo=TRUE, fig.align='center', fig.height=7, fig.width=7, warning=FALSE}
RES <- HT29R.expSimilarity(refDataDir = tmpDir, 
                           resDir = resultsDir,
                           geneGuides = "All",
                           geneLevel = FALSE,
                           Rscore = TRUE,
                           saveToFig = FALSE,
                           display = TRUE,
                           userFCs = UserData$logFCs)
```


Same as before but at the gene level.

```{r echo=TRUE, fig.align='center', fig.height=7, fig.width=7, warning=FALSE}
RES <- HT29R.expSimilarity(refDataDir = tmpDir, 
                           resDir = resultsDir,
                           geneGuides = "All",
                           geneLevel = TRUE,
                           Rscore = TRUE,
                           saveToFig = FALSE,
                           display = TRUE,
                           userFCs = UserData$logFCs)
```

## Phenotype intensity

Text

```{r echo=TRUE, fig.align='center', fig.height=10, fig.width=15, warning=FALSE}
layout(matrix(1, nrow=1, ncol=1, byrow=TRUE))

HT29R.PhenoIntensity(refDataDir = tmpDir,
                     resDir = resultsDir,
                     userFCs = UserData$logFCs, 
                     geneLevel = FALSE,
                     saveToFig = FALSE,
                     display = TRUE)

```

Same as before but at the gene level

```{r echo=TRUE, fig.align='center', fig.height=10, fig.width=15, warning=FALSE}
layout(matrix(1, nrow=1, ncol=1, byrow=TRUE))

HT29R.PhenoIntensity(refDataDir = tmpDir,
                     resDir = resultsDir,
                     userFCs = UserData$logFCs, 
                     geneLevel = TRUE,
                     saveToFig = FALSE,
                     display = TRUE)

```

## ROC analysis

Text

```{r echo=TRUE, fig.align='center', fig.height=5, fig.width=15, warning=FALSE}

# uncomment if geneLevel = TRUE
data("BAGEL_essential")
data("BAGEL_nonEssential")

Essential_sgRNAs <- ccr.genes2sgRNAs(KY_Library_v1.0, BAGEL_essential)
nonEssential_sgRNAs <- ccr.genes2sgRNAs(KY_Library_v1.0, BAGEL_nonEssential)

HT29R.ROCanalysis(refDataDir = tmpDir, 
                  positives = Essential_sgRNAs,
                  negatives = nonEssential_sgRNAs,
                  userFCs = UserData$logFCs,
                  geneLevel = FALSE,
                  saveToFig = FALSE,
                  display = TRUE)
```


Same as before, but at the gene level

```{r echo=TRUE, fig.align='center', fig.height=5, fig.width=15, warning=FALSE}

HT29R.ROCanalysis(refDataDir = tmpDir, 
                  positives = BAGEL_essential,
                  negatives = BAGEL_nonEssential,
                  userFCs = UserData$logFCs,
                  geneLevel = TRUE,
                  saveToFig = FALSE,
                  display = TRUE)
```


## HT-29-specifics fitness genes at 5% FDR and their characterisation

Text

```{r echo=TRUE, fig.align='center', fig.height=8, fig.width=6, warning=FALSE}

res <- HT29R.FDRconsensus(refDataDir = tmpDir,
                          resDir = resultsDir,
                          userFCs = UserData$logFCs,
                          distance = "Cohen's",
                          FDRth = 0.05,
                          saveToFig = FALSE,
                          display = TRUE)
```

Overlap with prior known (BAGEL) essential genes.

```{r echo=TRUE, warning=FALSE}
HT29R.FDRenrichment(consensus = res$POS, background=res$Universe, labels = BAGEL_essential)
```

## GO analysis

Top 10 Gene Ontology categories (Biological Process, BP) enriched for
the HT-29 Positive Consensus

```{r echo=TRUE, fig.height=5, fig.width=8, warning=FALSE}
BPmapping <- annFUN.org("BP", mapping = "org.Hs.eg.db", ID = "symbol") 
genesUniverse <- unique(unlist(BPmapping))

GO <- enrichGO(gene = res$POS, 
               keyType = "SYMBOL", 
               universe = genesUniverse, 
               ont="BP", 
               OrgDb = "org.Hs.eg.db")

dotplot(GO, showCategory=10)
```

## References
